### 2014-09-11 text/Chapter01.tex
Tweaks at the beginning and lofty closing paragraph (Chapter 1)

Many advances in programming language design are driven by some practical motivations. Sometimes,
the practical motivations are easy to see -- for example, when they come from an external change
such as the rise of multi-core processors. Sometimes, discovering the practical motivations is a
difficult task -- perhaps because we are so used to a certain way of doing things that we do not
even \emph{see} the flaws of our approach.

Before exploring the motivations leading to this thesis, we briefly consider two recent practical
concerns that led to the development of new programming languages. This helps to explain why
context-aware programming is of importance. The examples are by no means representative, but they
illustrate different kinds of motivations well.

@@

The rise of multi-core CPUs is a clear example of an external development influencing
programming language research. As multi-core and multi-processor systems became de-facto standard,
languages had to provide better abstractions for parallel programming. This led to the industrial
popularity of \emph{immutable} data structures (and functional programming in general), software
transactional memory \cite{app-haskell-stm}, data-parallelism and also asynchronous
computing \cite{app-fsharp-async}.

In this case, the motivation is easy to see -- writing multi-core programs using earlier
abstractions, such as threads and locks, is difficult and error-prone. At the same time,
multi-core CPUs become the standard very quickly and so the lack of good language abstractions
was apparent.

@@

Accessing data is an example of a more subtle challenge. Initiatives like open government data\footnote{
In the UK, the open government data portal is available at: \url{http://data.gov.uk/}}
certainly make more data available. However, to access the data, one has to parse CSV and Excel files,
issue SQL or SPARQL queries (to query database and the semantic web, respectively).

Technologies like LINQ \cite{app-linq} make querying data significantly easier. But perhaps
because accessing data became important more gradually, it was not easy to see that inline
SQL is a poor solution \emph{before} better approaches were developed.

This is even more the case for \emph{type providers} -- a recent feature in F\# that integrates
external data sources directly into the type system of the language and thus makes data explorable
directly from the source code editor (through features such as auto-completion on object members).
It is not easy to see the limitations of standard techniques (using HTTP requests to query REST
services or parsing CSV files and using string-based lookup) until one sees how type providers
change the data-scientist's workflow\footnote{This is difficult to explain in writing and so the
reader is encouraged to watch a video showing type providers for the WorldBank and CSV data
sources \cite{app-fsharp-world}.}.

@@

In this chapter, we argue that the next important practical challenge for programming language
designers is designing languages that are better at working with (and understanding) the
\emph{context in which programs are executed}.

This challenge is of the kind that is not easy to see -- perhaps because we are so used to
doing things in certain ways that we cannot see their flaws. In this chapter, we aim to
uncover such flaws -- we look at a number of basic programs that rely on contextual
information, we explain why they are inappropriate and then we briefly outline how this
thesis remedies the situation.

Putting deeper philosophical questions about the nature of scientific prog\-ress aside,
the goal of programming language research is generally to design languages that provide
more \emph{appropriate abstractions} for capturing common problems, are \emph{simple} and
more \emph{unified}.
These are exactly the aims that we follow in this thesis. In this chapter, we explain what
the common problems are. In Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural},
we develop two simple calculi to understand and capture the structure of the problems and,
finally, Chapter~\ref{ch:unified} unifies the two abstractions.

@@

The phrase \emph{context in which programs are executed} sounds rather abstract and generic.
What notions of \emph{context} can be identified in modern software systems?
Different environments provide different resources (\eg~database
or GPS sensor), environments are increasingly diverse (\eg~multiple versions of different
mobile platforms). Web applications are split between client, server and mobile components; mobile
applications must be aware of the physical environment while the ``internet of things'' makes
the environments even more heterogeneous. At the same time, applications access
rich data sources and need to be aware of security policies and provenance information from the
environment.

@@

\begin{itemize}
\item \textbf{System capabilities.} When writing code that is cross-compiled to multiple
  targets (e.g. SQL \cite{app-linq}, OpenCL or JavaScript \cite{app-hop-lang}) a part of the compilation
  (generating the SQL query) often occurs at runtime and developers have no guarantee that it will succeed
  until the program is executed.

@@

\item \textbf{Security and provenance.} When working with data (be it sensitive database or social
   network data), we have permissions to access only some of the data and we may want to track
   \emph{provenance} information. However, this is not checked -- if a program attempts to access
   unavailable data, the access will be refused at run-time.

@@

Most developers do not perceive the above as programming language flaws -- they are simply common
programming problems (at most somewhat annoying and tedious) that had to be solved. However, this
is because we do not realize that a suitable language extension could make the above problems
significantly easier to solve. As the number of distinct contexts and their diversity increases,
these problems will become even more commonplace.

The following sub-sections explore four examples in more details. The examples are chosen to
demonstrate two distinct forms of contexts that are studied in this thesis.

@@

Equally, in the .NET ecosystem, there are multiple versions of the .NET runtime, mobile and
portable versions of the framework etc. The differences may be subtle -- for example, some members
are omitted to make the mobile version of the library smaller, some functionality is not available
at all, but naming can also vary between versions.

@@

As an alternative to conditional compilation, developers can use dynamic loading. For example,
on Android, programs can access API from higher level platform dynamically using techniques like
reflection and writing wrappers. This is even more error prone. As noted in an article introducing
the technique\footnote{Retrieved from: \url{http://android-developers.blogspot.com/2009/04/backward-}\\\url{compatibility-for-android.html}}:
\emph{``Remember the mantra: if you haven't tried it, it doesn't work''}. Again, it would be
reasonable to expect that statically-typed languages could provide a better solution.

@@

Another example related to the previous one is when libraries use meta-programming techniques
(such as LINQ \cite{app-linq} or F\# quotations \cite{app-fsharp-metaprog}) to translate code
written in a subset of a host language to some other target language, such as SQL, OpenCL or
JavaScript. This is an important technique, because it lets developers targets multiple
heterogeneous runtimes that have limited execution capabilities.

For example, consider the following LINQ query written in C\# that queries a database and
selects product names where the first upper case letter is "C":
%
\begin{equation*}
\begin{array}{l}
\kvd{var}~\ident{db} = \kvd{new}~\ident{NorthwindDataContext}();\\[0.25em]
\kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
\kvd{where}~\ident{p.ProductName.First}(\ident{c} \Rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
\kvd{select}~\ident{p.ProductName};
\end{array}
\end{equation*}
%
This appears as a perfectly valid code ant the C\# compiler accepts it. However, when the program
is executed, it fails with the following error:
%
\begin{quote}
Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators
not supported for type \texttt{System.String}.
\end{quote}
%
The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
uses \ident{First} method to iterate over characters of a string, which is not supported. This is
not a technical limitation of LINQ, but a fundamental problem of the approach.

When cross-compiling to a limited environment, we cannot always support the full source language.
The example with LINQ and SQL demonstrates the importance of this problem. As of March 2014,
Google search returns 11800 results for the message above and even more (44100 results) for a
LINQ error message \emph{``Method X has no supported translation to SQL``} caused by a similar
limitation.

@@

The previous two examples were related to the non-existence of some library functions in another
environment. Another common factor was that they were related to the execution context of the whole
program or a scope. However, contextual properties can be also related to specific variables.

For example, consider the following code sample that accesses database by building a SQL query
using string concatenation:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{query} = \ident{sprintf}~\str{SELECT * FROM Products WHERE Name='\%s'}~\ident{name}\\[-0.25em]
\kvd{let}~\ident{cmd} = \kvd{new}~\ident{SqlCommand}(\ident{query})\\[-0.25em]
\kvd{let}~\ident{reader} = \ident{cmd.ExecuteReader}()
\end{array}
\end{equation*}
%
The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
(an attacker could enter \str{'; DROP TABLE Products -{}-} as their \ident{name} and delete the
database table with products). For this reason, most libraries discourage building SQL commands by
string concatenation, but there are still many systems that do so.

The example demonstrates a more general property. Sometimes, it is desirable to track additional
meta-data about variables that are in some ways special. Such meta-data can determine how the variables
can be used. Here, \ident{name} comes from the user input. This \emph{provenance} information should
be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
not directly contain user input (in an unchecked form). Such marking of values (but at run-time)
is also called tainting \cite{app-tainting-sql}.

@@

In another context, when working with data (\eg~in data journalism), it would be desirable to
track meta-data about the quality and the source of the data. For example, is the source trustworthy?
Is the data up-to-date? Such meta-data could propagate to the result and tell us important
information about the calculated results.

@@

The last example leaves the topic of cross-platform and distributed computing. We focus on
checking how arrays are accessed. This is a simpler version of the data-flow programming
examples used later in the thesis.

Consider a simple programming language with arrays where $n^{\textnormal{th}}$ element of an
array \ident{arr} is accessed using $\ident{arr}[n]$. Furthermore, we focus on performing local
transformations and we assume that the keyword $\kvd{cursor}$ returns the \emph{current} location
in the array.

The following example implements a simple one-dimensional cellular automata, reading from the
\ident{input} array and writing to \ident{output}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{sum}=\ident{input}[\kvd{cursor} - 1] ~+~ \ident{input}[\kvd{cursor}] ~+~ \ident{input}[\kvd{cursor} + 1]\\[-0.25em]
\kvd{if}~\ident{sum}=2 ~||~ (\ident{sum}=1 ~\&\&~ \ident{input}[\kvd{cursor} - 1] = 0)\\[-0.25em]
\kvd{then}~\ident{output}[\kvd{cursor}] \leftarrow 1 ~\kvd{else}~\ident{output}[\kvd{cursor}] \leftarrow 0
\end{array}
\end{equation*}
%
In this example, we use the term \emph{context} to refer to the values in the array around the
current location provided by \kvd{cursor}. The interesting question is, how much of the context
(\ie~how far in the array) does the program access.

This is contextual information attached to individual (array) variables.
In the above example, we want to track that \ident{input} is accessed in the range $\langle-1, 1 \rangle$
while \ident{output} is accessed in the range $\langle 0, 0 \rangle$. When calculating the ranges,
we need to be able to compose ranges $\langle-1,-1$, $\langle 0,0 \rangle$ and $\langle 1,1 \rangle$
(based on the accesses on the first line).

The information about access patterns can be used to efficiently compile the computation
(as we know which sub-range of the array might be accessed) and it also allows better handling
of boundaries. For example, wrap-around behaviour we could pad the \ident{input} with a known
number of elements from the other side of the array.

@@

The four examples presented in the previous section cover different notions of \emph{context}.
The context can be viewed as execution environment, capabilities provided by the environment
or input and meta-data about the input.

The different notions of context can be broadly classified into two categories -- those that
speak about the environment and those that speak about individual inputs (variables). In this
thesis, we refer to them as \emph{flat coeffects} and \emph{structural coeffects}, respectively:

\begin{itemize}
\item \textbf{Flat coeffects} represent additional data, resources and meta-data that are
   available in the execution environment (regardless of how they are accessed in a program).
   Examples include resources such as GPS sensors and battery status (on a phone), databases
   (on the server), or framework version.

@@

\subsection{Context-aware language in action}
\label{sec:intro-context-example}

As an example, consider a news reader consisting of server-side (which stores the news in a
database) and a number of clients applications for popular platforms (Android, Windows Phone, etc.).
A simplified code excerpt that might appear somewhere in the implementation is shown in
Figure~\ref{fig:introduction-sample}.

@@

If we were writing the code using current main-stream technologies, we would have to create three
completely separate components. The server-side would include the \ident{fetchNews} function, which
queries the database. The iPhone version would include \ident{fetchLocalNews}, which gets the current
GPS location and performs a call to the remote server and \ident{iPhoneMain}, which constructs the
user-interface. For Windows, we would also need \ident{fetchLocalNews}, but this time with
\ident{windowsMain}. When using a language that can be compiled for all of the platforms, we would need
a number of \prepk{\#if} blocks to delimit the platform-specific parts.

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{fetchNews}(\ident{loc}) =\\[-0.25em]
\hspace{1em} \kvd{let}~\ident{cmd} = \ident{sprintf}~\str{SELECT * FROM News WHERE Location='\%s'}~\ident{loc}\\[-0.25em]
\hspace{1em} \ident{query}(\ident{cmd}, \ident{password})\\[-0.25em]
\\[-1.0em]
\kvd{let}~\ident{fetchLocalNews}() =\\[-0.25em]
\hspace{1em} \kvd{let}~\ident{loc} = \ident{gpsLocation}()\\[-0.25em]
\hspace{1em} \kvd{remote}~\ident{fetchNews}(\ident{loc})\\[-0.25em]
\\[-1.0em]
\kvd{let}~\ident{iPhoneMain}() = \\[-0.25em]
\hspace{1em} \ident{createCocoaListing}(\ident{fetchLocalNews})\\[-0.25em]
\\[-1.0em]
\kvd{let}~\ident{windowsMain}() = \\[-0.25em]
\hspace{1em} \ident{createMetroListing}(\ident{fetchLocalNews})
\end{array}
\end{equation*}

@@

Here, we use the notation $\tau_1 \,\ident{\hspace{-1em}\footnotesize{@}}\, c \rightarrow \tau_2$ to denote
a function that takes an input of type $\tau_1$, produces a result of type $\tau_2$ and has additional
context requirements specified by $c$. In the above example, the annotation $c$ is simply a set of
required resources or capabilities. However, a more complex structure could be used as well, for example,
including the Android API level.

The following summary shows the types of the functions from the code sample in
Figure~\ref{fig:introduction-sample}. These guide the code generation by specifying which
function should be compiled for which of the platforms, but they also provide documentation
for the developers:

\begin{equation*}
\begin{array}{lcl}
\ident{password} &:& $\introvar{string}{sensitive}$\\[-0.25em]
\ident{fetchNews} &:& $\introfunc{location}{database}{news~list}$\\[0.5em]
\ident{gpsLocation} &:& $\introfunc{unit}{gps}{location}$\\[-0.25em]
\ident{fetchLocalNews} &:& $\introfunc{location}{gps, rpc}{news~list}$\\[0.5em]
\ident{iPhoneMain} &:& $\introfunc{unit}{ cocoa, gps, rpc}{unit}$\\[-0.25em]
\ident{windowsMain} &:& $\introfunc{unit}{ windows, gps, rpc}{unit}$
\end{array}
\end{equation*}

As mentioned earlier, the concrete syntax used here is just for illustration. Furthermore,
some information could even be mapped to other visual representations -- for example, differently
coloured backgrounds for platform-specific functions. The key point is that the type provides a
number of useful information:

@@

\item We can compile the program for two client-side platforms - the entry points are \ident{iPhoneMain}
  and \ident{windowsMain} and require Cocoa and Windows user-interface libraries, together with
  GPS and the ability to perform remote calls over the network.
\end{itemize}

@@

The previous section introduced the idea of context-aware languages from the practical perspective.
As already discussed, we approach the problem from the perspective of statically typed programming
languages. This section outlines how can contextual information be integrated into the
standard framework of static typing. This section is intended only as an informal overview and the
related work is discussed in Chapter~\ref{ch:pathways}.

\paragraph{Type systems.}
A type system is a form of static analysis that is usually specified by \emph{typing judgements}
such as $\Gamma \vdash e : \tau$. The judgement specifies that, given some variables described by
the context $\Gamma$, the expression $e$ has a type $\tau$.
The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an
expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int}
or \ident{float}, depending on the types of the variables. For example, the following is a valid
typing judgement in F\#:
%
\begin{equation*}
x\!:\!\ident{int},~y\!:\!\ident{int} \vdash x+y : \ident{int}
\end{equation*}
%
This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
be \ident{int}. The expression would also be typeable in a context $x\!:\!\ident{int},~y\!:\!\ident{int}$,
but not, for example, in a context where $x$ has a type \ident{unit}.

>>>>

Many advances in programming language design are driven by practical motivations. Sometimes,
these practical motivations are easy to see -- for example, when they come from an external change
such as the rise of multi-core processors. Sometimes, discovering the practical motivations is a
difficult task -- perhaps because we are so used to a certain way of doing things that we do not
even \emph{see} the flaws of our approach.

Before exploring the motivations for to this thesis, we briefly consider two recent practical
concerns that have led to the development of new programming languages. This helps to explain why
context-aware programming is important. The examples are by no means representative, but they
illustrate various kinds of motivations well.

@@

The rise of multi-core CPUs is a clear example of an external development influencing
programming language research. As multi-core and multi-processor systems became ubiquitous,
languages had to provide better abstractions for parallel programming. This led to the industrial
popularity of \emph{immutable} data structures (and functional programming in general), software
transactional memory \cite{app-haskell-stm}, data-parallelism and also asynchronous
computing \cite{app-fsharp-async}.

In this case, the motivation is easy to see -- writing multi-core programs using earlier
abstractions, such as threads and locks, is difficult and error-prone. At the same time,
multi-core CPUs became a standard very quickly and so the lack of good language abstractions
was apparent.

@@

Accessing ``big data'' sources is an example of a more subtle challenge. Initiatives like open 
government data\footnote{In the UK, the open government data portal is available at: \url{http://data.gov.uk/}}
certainly make more data available. However, to access the data, one has to parse CSV and Excel files,
issue SQL or SPARQL queries (to query database and the semantic web, respectively).

Technologies like LINQ \cite{app-linq} make querying data significantly easier. But perhaps
because accessing data became important more gradually, it was not easy to see that SQL queries,
embedded as parameterized strings\footnote{The dominant approach is demonstrated, for example, 
by a review of SQL injection prevention techniques by Clarke \cite{app-sql-injection}}, are a 
poor solution \emph{before} better approaches were developed.

This is even more the case for \emph{type providers} -- a recent feature in F\# that integrates
external data sources directly into the type system of the language and thus makes data explorable
directly from the source code editor (through features such as auto-completion on object members).
It is not easy to see the limitations of standard techniques (using HTTP requests to query REST
services or parsing CSV files and using string-based lookup) until one sees just how much type 
providers change the data-scientist's workflow\footnote{This is difficult to explain in writing 
and so the reader is encouraged to watch a video showing type providers for the WorldBank and CSV 
data sources \cite{app-fsharp-world}.}.

~\\

@@

In this thesis, we argue that the next important practical challenge for programming language
designers is designing languages that are better at working with (and understanding) the
\emph{context in which programs are executed}.

This challenge is of the kind that is not easy to see, perhaps because we are so used to
doing things in certain ways that we cannot see their flaws. In this chapter, we aim to
expose such flaws. We look at a number of basic programs that rely on contextual
information, we explain why the currently dominant solutions are inappropriate and then 
briefly outline how this thesis solves the problems.

Putting deeper philosophical questions about the nature of scientific prog\-ress aside,
the goal of programming language research is generally to design languages that provide
more \emph{appropriate abstractions} for capturing common problems, are \emph{simple} and
more \emph{unified}. These are exactly the aims that we follow in this thesis. In this chapter, 
we explain what the common problems in context-dependent programming are. In 
Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}, we develop two simple calculi to 
understand and capture the structure of those problems and, finally, Chapter~\ref{ch:unified} 
unifies the two abstractions.

% ==================================================================================================
%                                                                                             
%     ###                  #                    #                                              
%    #   #                 #                    #                                              
%    #       ###   # ##   ####    ###   #   #  ####           ###   #   #   ###   # ##    ###  
%    #      #   #  ##  #   #     #   #   # #    #     #####      #  #   #      #  ##  #  #   # 
%    #      #   #  #   #   #     #####    #     #             ####  # # #   ####  #      ##### 
%    #   #  #   #  #   #   #  #  #       # #    #  #         #   #  # # #  #   #  #      #     
%     ###    ###   #   #    ##    ###   #   #    ##           ####   # #    ####  #       ###  
%                                                                                                                                                                       % ==================================================================================================

@@

The phrase \emph{context in which programs are executed} sounds rather abstract and generic.
What notions of \emph{context} can be identified in modern software systems?
Different environments provide different resources (\eg~a database or GPS sensors), environments 
are increasingly diverse (\eg~different mobile platforms with multiple partially incompatible versions). 
Web applications are split between client, server and mobile components; mobile
applications must be aware of the physical environment while the ``internet of things'' makes
the environment even more heterogeneous. At the same time, applications access
rich data sources and need to be aware of provenance information and respect the security policies 
from the environment.

@@

\begin{itemize}
\item \textbf{System capabilities.} Libraries such as LINQ \cite{app-linq} let developers write code
  in a host language like C\# and then cross-compile it to multiple targets (including SQL, OpenCL 
  or JavaScript \cite{app-hop-lang}). Part of the compilation (\eg~generating the SQL query) occurs 
  at runtime and developers have no guarantee that it will succeed until the program is executed,
  because only subset of the host language is supported.

@@

\item \textbf{Security and provenance.} When working with data (be it sensitive database or social
   network data), we may have permission to access only some of the data and we may want to track
   \emph{provenance} information. However, this is not checked -- if a program attempts to access
   unavailable data, the access will be refused at run-time.

@@

\noindent
Most developers do not perceive the above as programming language flaws -- they are simply common
programming problems (at most somewhat annoying and tedious) that have to be solved. However, this
is because it is not apparent that a suitable language extension could make the above problems
significantly easier to solve. As the number of distinct contexts and their diversity increases,
these problems will become even more commonplace.

The following sub-sections explore 4 examples in more detail. The examples are chosen to
demonstrate two distinct forms of contexts that are studied in this thesis -- first two are
related to the program environment and the latter two are associated with individual variables of 
the program.

@@

Equally, in the .NET ecosystem, there are multiple versions of the .NET runtime, mobile and
portable versions of the framework etc. The differences may be subtle -- for example, some instance
methods and properties are omitted to make the mobile version of the library smaller, some functionality 
is not available at all, but naming can also vary between versions.

@@

As an alternative to conditional compilation, developers can use dynamic loading. For example,
on Android, programs can access API from higher level platform dynamically using techniques like
reflection and writing wrappers. This is even more error prone. As noted in an article\footnote{Retrieved 
from: \url{http://android-developers.blogspot.com/2009/04/backward-}\\\url{compatibility-for-android.html}}
introducing the technique ``Remember the mantra: if you haven't tried it, it doesn't work.'' 
Again, it would be reasonable to expect that statically-typed languages could provide a better solution.

@@

Another example related to the previous one is when libraries use meta-programming techniques
(such as LINQ \cite{app-linq} or F\# quotations \cite{app-fsharp-metaprog}) to translate code
written in a subset of a host language to some other target language, such as SQL, OpenCL or
JavaScript. For database access, this is a recently developed technique replacing embedded SQL
discussed in the introduction, but it is a more generally important technique for programming
in heterogeneous environments. It lets developers targets multiple runtimes that have limited 
execution capabilities.

For example, the following LINQ query written in C\# queries a database and
selects those product names where the first upper case letter is ``C'':
%
\begin{equation*}
\begin{array}{l}
\kvd{var}~\ident{db} = \kvd{new}~\ident{NorthwindDataContext}();\\[0.25em]
\kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
\kvd{where}~\ident{p.ProductName.First}(\lambda \ident{c} \rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
\kvd{select}~\ident{p.ProductName};
\end{array}
\end{equation*}
%
This appears as a perfectly valid code and the C\# compiler accepts it. However, when the program
is executed, it fails with the following error:
%
\begin{quote}
Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators
not supported for type \texttt{System.String}.
\end{quote}
%
The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
uses the \ident{First} method to iterate over characters of a string, which is not supported. This is
not a technical limitation of LINQ, but a fundamental problem of the approach.

When cross-compiling to a limited environment, we cannot always support the full source language.
The example with LINQ and SQL demonstrates the importance of this problem. As of March 2014,
Google search returns 11800 results for the message above and even more results (44100) for a
LINQ error message \emph{``Method X has no supported translation to SQL''} caused by a similar
limitation.

@@

The previous two examples were related to the non-existence of some library functions in a different
execution environment. Another common factor was that they were related to the execution context of 
the whole program or a function scope. However, contextual properties can also be associated with
a specific variables.

For example, consider the following code sample that accesses a database by building a SQL query
using string concatenation. For the purpose of the demonstration, this example does not use LINQ, 
but an older approach with a parameterized SQL query written as a string:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{query} = \ident{sprintf}~\str{SELECT * FROM Products WHERE Name='\%s'}~\ident{name}\\[-0.25em]
\kvd{let}~\ident{cmd} = \kvd{new}~\ident{SqlCommand}(\ident{query})\\[-0.25em]
\kvd{let}~\ident{reader} = \ident{cmd.ExecuteReader}()
\end{array}
\end{equation*}
%
The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
\cite{app-sql-injection} (an attacker could enter \str{'; DROP TABLE Products -{}-} as their 
\ident{name} and delete the database table ``Products''). For this reason, most libraries discourage 
building SQL commands by string concatenation, but there are still many systems that do so.

The example demonstrates a more general property. Sometimes, it is desirable to track additional
meta-data about variables that are in some ways special. Such meta-data can determine how the variables
can be used. Here, \ident{name} comes from the user input. This information about the value should
be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
not directly contain user input (in an unchecked form).

@@

In the security context, such marking of values (but at run-time) is called \emph{tainting} 
\cite{app-tainting-sql}, but the technique is a special case of more general \emph{provenance}
tracking. This can be useful when working with data in other contexts. For example, 
data jounralsts might want to propagate meta-data about the quality and the information source -- 
is the source trustworthy? Is the data up-to-date? Such meta-data could propagate to 
the result and tell us important information about the calculated results.

@@

The final example leaves the topic of cross-platform and distributed computing. We focus on
checking how arrays are accessed. This is a simpler version of the data-flow programming
examples used later in the thesis.

Consider a simple programming language with arrays where $n^{\mathit{th}}$ element of an
array \ident{arr} is accessed using $\ident{arr}[n]$. We focus on writing stencil computations
(such as image blurring, Conway's game of life or convolution) where all arrays are of
the same size and the system provides a \emph{cursor} pointing to a current location in the 
stencil. We assume that the keyword \kvd{cursor} returns the current location in the stencil.

The following example implements a simple one-dimensional cellular automaton, reading from the
\ident{input} array and writing to \ident{output}:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{sum}=\ident{input}[\kvd{cursor} - 1] ~+~ \ident{input}[\kvd{cursor}] ~+~ \ident{input}[\kvd{cursor} + 1]\\[-0.25em]
\kvd{if}~\ident{sum}=2 ~||~ (\ident{sum}=1 ~\&\&~ \ident{input}[\kvd{cursor} - 1] = 0)\\[-0.25em]
\kvd{then}~\ident{output}[\kvd{cursor}] \leftarrow 1 ~\kvd{else}~\ident{output}[\kvd{cursor}] \leftarrow 0
\end{array}
\end{equation*}
%
In this example, we use the term \emph{context} to refer to the values in the array around the
current location provided by \kvd{cursor}. The interesting question is, how much of the context
(\ie~how far in the array) does the program access.

This is contextual information attached to individual (array) variables.
In the above example, we want to track that \ident{input} is accessed in the range $\langle-1, 1 \rangle$
while \ident{output} is accessed in the range $\langle 0, 0 \rangle$. When calculating the ranges,
we need to be able to compose ranges $\langle-1,-1 \rangle$, $\langle 0,0 \rangle$ and $\langle 1,1 \rangle$
(based on the three accesses on the first line).

The information about access patterns can be used to efficiently compile the computation
by preallocating the necessary space (as we know which sub-range of the array might be accessed).
It also allows better handling of boundaries. For example, to simplify wrap-around behaviour we 
could pad the \ident{input} with a known number of elements from the other side of the array.

% ==================================================================================================
%
%    #                                                             
%    #                                                             
%    #       ###   # ##    ## #  #   #   ###    ## #   ###    ###  
%    #          #  ##  #  #  #   #   #      #  #  #   #   #  #     
%    #       ####  #   #   ##    #   #   ####   ##    #####   ###  
%    #      #   #  #   #  #      #  ##  #   #  #      #          # 
%    #####   ####  #   #   ###    ## #   ####   ###    ###   ####  
%                         #   #                #   #               
%                          ###                  ###                
%
% ==================================================================================================

@@

The four examples presented in the previous section cover different kinds of \emph{context}.
The context includes notions such as execution environment, capabilities provided by the 
environment or input and meta-data about the input.

The different applications can be broadly classified into two categories -- those that
speak about the environment and those that speak about individual inputs (variables). In this
thesis, we refer to them as \emph{flat coeffects} and \emph{structural coeffects}, respectively:

\begin{itemize}
\item \textbf{Flat coeffects} represent additional data, resources and meta-data that are
   available in the execution environment (regardless of how they are accessed in a program).
   Examples include resources such as GPS sensors and battery status (on a phone), databases
   (on the server), or software framework (or library) version.

@@

\subsection{Context-aware languages in action}
\label{sec:intro-context-example}

As an example, consider a news reader app consisting of a server-side component (which stores the 
news in an SQL database) and a number of clients applications for popular platforms (Android, Windows 
Phone, etc.). A simplified code excerpt that might appear somewhere in the implementation is shown in
Figure~\ref{fig:introduction-sample}.

@@

If we were writing the code using current mainstream technologies, we would have to create three
completely separate components. The server-side would include the \ident{fetchNews} function, which
queries the database. The iPhone version would include \ident{fetchLocalNews}, which gets the current
GPS location and performs a call to the remote server and \ident{iPhoneMain}, which constructs the
user-interface. For Windows, we would also need \ident{fetchLocalNews}, but this time with
\ident{windowsMain}. When using a language that can be compiled for all of the platforms, we would need
a number of \prepk{\#if} blocks to delimit the platform-specific parts.

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{fetchNews}(\ident{loc}) =\\[-0.25em]
\hspace{1em} \kvd{let}~\ident{cmd} = \ident{sprintf}~\str{SELECT * FROM News WHERE Location='\%s'}~\ident{loc}\\[-0.25em]
\hspace{1em} \ident{query}(\ident{cmd}, \ident{password})\\[-0.25em]
\\[-1.0em]
\kvd{let}~\ident{fetchLocalNews}() =\\[-0.25em]
\hspace{1em} \kvd{let}~\ident{loc} = \ident{gpsLocation}()\\[-0.25em]
\hspace{1em} \kvd{remote}~\ident{fetchNews}(\ident{loc})\\[-0.25em]
\\[-1.0em]
\kvd{let}~\ident{iPhoneMain}() = \\[-0.25em]
\hspace{1em} \ident{createiPhoneListing}(\ident{fetchLocalNews})\\[-0.25em]
\\[-1.0em]
\kvd{let}~\ident{windowsMain}() = \\[-0.25em]
\hspace{1em} \ident{createWindowsListing}(\ident{fetchLocalNews})
\end{array}
\end{equation*}

@@

\noindent
Here, we use the notation $\tau_1 \,\ident{\hspace{-1em}\footnotesize{@}}\, c \rightarrow \tau_2$ to denote
a function that takes an input of type $\tau_1$, produces a result of type $\tau_2$ and has additional
context requirements specified by $c$. In the above example, the annotation $c$ is simply a set of
required resources or capabilities. However, a more complex structure could be used as well, for example,
including the Android API level as an integer.

The following summary shows the types of the functions from the code sample in
Figure~\ref{fig:introduction-sample}. These guide code generation by specifying which
function should be compiled for which of the platforms, but they also provide documentation
for the developers. In addition to function annotations, we also show the annotation attached
to the \ident{password} variable:
%
\begin{equation*}
\begin{array}{lcl}
 \ident{password} &:& $\introvar{string}{sensitive}$\\[-0.25em]
 \ident{fetchNews} &:& $\introfunc{location}{database}{news~list}$\\[0.5em]
 \ident{gpsLocation} &:& $\introfunc{unit}{gps}{location}$\\[-0.25em]
 \ident{fetchLocalNews} &:& $\introfunc{location}{gps, rpc}{news~list}$\\[0.5em]
 \ident{iPhoneMain} &:& $\introfunc{unit}{ ios, gps, rpc}{unit}$\\[-0.25em]
 \ident{windowsMain} &:& $\introfunc{unit}{ windows, gps, rpc}{unit}$
\end{array}
\end{equation*}
%
The example combines two separate notions of context. The variable \ident{password} is annotated 
with a single (per-variable) annotation specifying tainting while
functions are annotated with a set of resource requirements.

The concrete syntax used here is just for illustration. Furthermore,
some information could even be mapped to other visual representations -- for example, differently
coloured backgrounds for platform-specific functions. The key point is that the type provides a
number of useful information:

@@

\item We can compile the program for two client-side platforms - the entry points are \ident{iPhoneMain}
  and \ident{windowsMain} and require iOS and Windows user-interface libraries, together with
  GPS and the ability to perform remote calls over the network.
\end{itemize}

@@

The previous section introduced the idea of context-aware languages from the practical perspective.
As already discussed, we approach the problem from the perspective of statically typed programming
languages. This section outlines how can contextual information be integrated into the
standard framework of static typing. This section is intended only as an informal overview and 
complete review of related work is available in Chapter~\ref{ch:pathways}.

\paragraph{Type systems.}
A type system is a form of static analysis that is usually specified by \emph{typing judgements}
such as $\Gamma \vdash e : \tau$. The judgement specifies that, given some variables described by
the context $\Gamma$, the expression $e$ has a type $\tau$.
The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an
expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int},
\ident{float}, or even \ident{string}, depending on the types of the variables. For example, 
the following is a valid typing judgement in F\#:
%
\begin{equation*}
x\!:\!\ident{int},~y\!:\!\ident{int} \vdash x+y : \ident{int}
\end{equation*}
%
This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
be \ident{int}. In F\#, the expression would also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
but not, for example, in a context where $x$ has a type \ident{int} and $y$ has a type \ident{string}.
